Categorical Foundations of E-Graphs With Binding

### 2025-07-09 POPL2026/notes/SlottedNotes.tex
comparison, semantics, enrichment fix, appendix

\begin{table}
\begin{tabular}{lcccc}
  Set of equations & built-in subst & explicit subst & explicit subst & explicit subst\\
  &&& + interchange&  + interchange\\
  &&&&  + comm\\
  \# of e-nodes & 40 & 161 & 289 & 356
\end{tabular}
\caption{Slotted e-graphs benchmarks}
\label{tbl:slotted}
\end{table}

@@

>>>>

\begin{table}
\begin{tabular}{lccccc}
  Set of equations & built-in subst & explicit subst & explicit subst & explicit subst & explicit subst\\
  &&& + comm& + interchange&  + interchange\\
  &&&&&  + comm\\
  \# of e-nodes & 40 & 161 & 288 & 289 & 356
\end{tabular}
\caption{Slotted e-graphs benchmarks}
\label{tbl:slotted}
\end{table}

@@

\begin{remark}

The rules arise from linear substitution lambda calculus which is as follows

\begin{align*}
(\lambda\;x\;t)\;L\;u \mapsto_{db} t[x / u]L\\
C\llbracket x \rrbracket [x / u] \mapsto_{ls} C \llbracket u \rrbracket [x / u]\\
t[x/u] \mapsto t, \text{if } x \not \in fv(t)
\end{align*}

The contexts $C$ are defined as

\[
C \Coloneqq \square \;|\; \lambda x . C \; | \; C\;t \;|\; t\;C\;|\;C[x / t]\;|\;t[x / C]
\]

and in particular in $\lambda x . C$ and $C[x/t]$ $x$ is bound.

\end{remark}

### 2025-07-09 POPL2026/paper/sections/categorical_semantics.tex
comparison, semantics, enrichment fix, appendix

    \begin{proposition}(A specialised case of Proposition 6.4.7~\cite{Borceux_1994})
        There is a 2-adjunction 
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
        % \[\begin{tikzcd}
        % 	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        \mathcal{F} \dashv \mathcal{U} : \catname{SLat}\text{--}\catname{Cat} \to \catname{Cat}
        \vspace{-1mm}
        \]
        that is induced by the usual free-forgetful adjunction

        % \[\begin{tikzcd}
        % 	\catname{SLatt} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        F \dashv U : \catname{SLat} \to \catname{Set}
        \vspace{-1mm}
        \]        
    \end{proposition}
In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor, it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{Slat}$-adjunction.
Thus, applying the functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{Slat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{equation}
\begin{aligned}
\Lambda_{A,B,C}(f + g) &= \eta;(B \multimap (f + g))\\
                       &= \eta;(B \multimap (f) + B \multimap (g))\\
                       &= \eta;(B \multimap (f)) + \eta;(B \multimap (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{aligned}
\end{equation}
\label{law:distributivity}

@@

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Fig. \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.
Note the orientation of string diagrams are now from top to bottom as opposed to what was used in~\nameref{sec:introduction}: the latter was mimicking the orientation of e-graphs, \textit{i.e.}, traditional orientation for expression trees.
% Fig. \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
% The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
% We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

@@

\begin{figure*}
\[
\adjustbox{scale=0.65}{
\tikzfig{./figures/egraph-strings}
}
\]
\captionsetup{skip=0pt, belowskip=-2ex}
\caption{String diagrams for closed semilattice-enriched symmetric monoidal categories.}
\label{fig:egraph-strings}
\end{figure*}

>>>>

    \begin{proposition}(A specialised case of Proposition 6.4.7~\cite{Borceux_1994} and Theorem 5.7.1~\cite{cruttwell2008normed})
        There is a 2-adjunction 
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
        % \[\begin{tikzcd}
        % 	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        \mathcal{F} \dashv \mathcal{U} : \catname{SLat}\text{--}\catname{Cat} \to \catname{Cat}
        \vspace{-1mm}
        \]
        that is induced by the usual (symmetric monoidal) free-forgetful adjunction

        % \[\begin{tikzcd}
        % 	\catname{SLatt} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        F \dashv U : \catname{SLat} \to \catname{Set}.
        \vspace{-1mm}
        \]
        Moreover, every symmetric monoidal category $\mathbb{C}$ in $\catname{Cat}$ (respectively, $\catname{SLat}\text{--}\catname{Cat}$) is taken to a symmetric monoidal category $\mathbb{C}'$ in $\catname{SLat}\text{--}\catname{Cat}$ (respectively, $\catname{Cat}$) by a 2-functor $\mathcal{F}$ (respectively, $\mathcal{U}$).
    \end{proposition}
In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor, it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{SLat}$-adjunction.
Thus, applying the 2-functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{SLat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{equation}
\begin{aligned}
\Lambda_{A,B,C}(f + g) &= \eta;(B \multimap (f + g))\\
                       &= \eta;(B \multimap (f) + B \multimap (g))\\
                       &= \eta;(B \multimap (f)) + \eta;(B \multimap (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{aligned}
\end{equation}
\label{law:distributivity}

@@

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Fig. \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.

% rotated the diagrams
% Note the orientation of string diagrams are now from top to bottom as opposed to what was used in~\nameref{sec:introduction}: the latter was mimicking the orientation of e-graphs, \textit{i.e.}, traditional orientation for expression trees.

% Fig. \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
% The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
% We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

@@

\begin{figure*}
\[
\adjustbox{scale=0.65}{
\tikzfig{./figures/egraph-strings-2}
}
\]
\captionsetup{skip=0pt, belowskip=-2ex}
\caption{String diagrams for closed semilattice-enriched symmetric monoidal categories.}
\label{fig:egraph-strings}
\end{figure*}

\subsection{Categorical foundations for slotted e-graphs}

Slotted e-graphs~\cite{slotted-egraphs} extended the traditional signature employed by e-graphs by an explicit construct for binding ($\text{bind}\; \$x\; tc$) and variables (slots).
\begin{align*}
&\text{functional symbols}\; f, g, \ldots\\
&\text{slots}\;  \$x, \$y, \ldots\\
&\text{term children}\; tc  \Coloneqq t \;|\; tc_{1}, \ldots, tc_{n} \;|\; \text{bind}\; \$x\; tc \;|\; \$x\\ 
&\text{terms}\;  \Coloneqq f \;|\; f(tc)
\end{align*}

The traditional congruence relation maintained be e-graphs for a set of equations $E$ that are given below (plus obvious rules for transitivity, symmetry and reflexivity of $\cong$)

\begin{mathpar}
\inferrule*[right=start]
    {a = b \in E}
    {E \vdash a \cong b}
    \inferrule*[right=cong. (variadic)]
      {E \vdash tc_i \cong tc'_i,\; i = 1, \ldots, k}
      {E \vdash tc_1, \ldots, tc_k \cong tc'_1, \ldots, tc'_k}
    \inferrule*[right=cong. (f)]
    {E \vdash tc \cong tc'}
    {E \vdash f(tc_1) \cong f(tc')}
\end{mathpar}

\noindent
are extended by the rule

\begin{mathpar}

\inferrule*[right=cong. (bind)]
    {E \vdash tc \cong tc'}
    {E \vdash \text{bind}\; \$x\; tc \cong \text{bind}\; \$x\; tc'},
\end{mathpar}

\noindent
plus the rules for equivalence of terms up to renaming of slots.
Recalling that we use $+$ to encode this congruence of terms, the first three rules (plus reflexivity, symmetry, and transitivity) correspond to idempotence, commutativity and associativity of $+$ and the fact that $+$ respect the composition $(;)$.
%  and tensoring $\otimes$ (traditional e-graphs do not have tensoring as they are defined for algebraic signatures).
While the other rule is \textit{natural} to enforce when developing e-graphs with binding, we see that it is indeed the \textit{right} rule to enforce as we can see it follow by a purely categorical argument by the free enrichment as per the law \ref{law:distributivity}.
We do not enforce any rules for renaming as string diagrams are essentially a nameless representation of terms.
Moreover, some renaming rules are not consistent with string diagrams, as slotted e-graph try to represent, \textit{e.g.}, terms $(a + b)$ and $(c + d)$ as a single entity (up to renaming of free variables), which is not the case for string diagrams (unless we do closure conversion).

% The above distributivity law closely correspond to one of the congruences maintained by a slotted e-graph, namely

% \[
% \text{bind}\;s_1\; x \cong \text{bind}\;s_2\;y \;\; \text{if}\; x \cong_{\alpha} y
% \]
% The congruence intuitively says that if $x$ and $y$ are congruent (up to $\alpha$-equivalence), then the enclosing binders are also congruent.
% While this is a \textit{natural} congruence to define when constructing an e-graph with support for binders, in our framework it follows from the free enrichment by a purely categorical argument justifying that it is the \textit{right} congruence to define.

### 2025-07-09 POPL2026/paper/sections/introduction.tex
mostly finish lambda calculus section
comparison, semantics, enrichment fix, appendix

\subsection{Conventional vs. categorical e-graphs with binding}

@@

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition. 
This is appears to be simpler just by comparing diagrams, but the technical advantages of using string diagrams run deep.
The explicit occurrences of variable names in the older approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices when performing $\beta$-reduction would hardly solve this problem, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.
A recent work of~\cite{slotted-egraphs} on so-called \textit{slotted} e-graphs has addressed the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by making variables into first-class citizens of an e-graph and made binding a built-in operation for an e-graph.
An example of a slotted e-graph that represents a term $\lambda x . (a + b) + (x + c)$ can be seen in Figure~\ref{fig:slotted}(left).

@@

A notable feature of a slotted e-graph is that every e-class is parameterised by a set of free variables of the terms represented by this e-class.
The references then additionally convey the information about how variables are instantiated.
For example, an e-class containing $+$ that is parameterised by variable names $s_2$ and $s_3$ gets intantiated once with $a,b$ (via $s_4 \mapsto a, s_2 \mapsto s_4$ and $s_5 \mapsto b, s_3 \mapsto s_5$) and once  with $x,c$.
The fact that the variable $x$ is bound is evident from the corresponding e-class being parameterised by only three variables.
If we were to add another e-node such as $e + d$, there would be a reference added to an already existing e-class such that $s_1 \mapsto e$, $s_2 \mapsto d$.
We can achieve the same degree of sharing (in the sense we have the same number of nodes for $+$) using string diagrams by doing a closure conversion of $+$ (Figure~\ref{fig:slotted}(right)).
However, the addition of these parameters and the corresponding alternations in the algorithmics makes the deeper correspondence between string diagrams and slotted e-graphs obscure.
Despite this, there are applications where the use of string diagrams may be more beneficial than that of (slotted) e-graphs, for example in the case of lambda calculus with explicit substitution that we will describe next.
Also, the categorical model that we built can help to explain the congruences maintained by slotted e-graphs which is further elaborated on in Section~\ref{sec:categorical}.

@@

\subsection{Lambda calculus with explicit substitutions and graphical equivalence of terms}

>>>>

\subsection{Conventional vs.\ categorical e-graphs with binding}%
\label{sec:vs-e-graphs-with-binding}

@@

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition. 
This is appears to be simpler just by comparing diagrams, but the technical advantages of using string diagrams run deep.
The explicit occurrences of variable names in the older approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices when performing $\beta$-reduction would hardly solve this problem, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.
A recent work of~\cite{slotted-egraphs} on so-called \textit{slotted} e-graphs has addressed the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by making variables into first-class citizens of an e-graph and made binding a built-in operation for an e-graph.
They also solved the problem of explicit variable names guaranteeing freshness inside binders explicitly differentiating between bound and free variables.
An example of a slotted e-graph that represents a term $\lambda x . (a + b) + (x + c)$ can be seen in Figure~\ref{fig:slotted} (left).

@@

A notable feature of a slotted e-graph is that every e-class is parameterised by a set of free variables of the terms represented by this e-class.
The references then additionally convey the information about how variables are instantiated.
For example, an e-class containing $+$ that is parameterised by variable names $s_2$ and $s_3$ gets intantiated once with $a,b$ (via $s_4 \mapsto a, s_2 \mapsto s_4$ and $s_5 \mapsto b, s_3 \mapsto s_5$) and once  with $x,c$.
The fact that the variable $x$ is bound is evident from the corresponding e-class being parameterised by only three variables.
If we were to add another e-node such as $e + d$, there would be a reference added to an already existing e-class such that $s_1 \mapsto e$, $s_2 \mapsto d$.
We can achieve the same degree of sharing (in the sense that we have the same number of nodes for $+$) using string diagrams by performing a closure conversion of $+$ (Figure~\ref{fig:slotted} (right)).
However, the addition of these parameters and the corresponding alterations in the algorithmics makes the deeper correspondence between string diagrams and slotted e-graphs obscure.
In fact, slotted e-graph appears to be a more general construction in a sense that it supports equality saturation for any signature with binders, such as, for example, the $\pi$-calculus, while our approach is tailored towards structures that arise from (Cartesian) closed symmetric monoidal categories.
Therefore, the use of string diagrams may be more advantageous than that of (slotted) e-graphs in the case of the latter.
We defer a case study to Section~\ref{sec:application}.
Finally, the categorical model that we develop can help justify the congruences maintained by slotted e-graphs which is further elaborated on in Section~\ref{sec:categorical}.

@@

% Do we need this section?
% \subsection{Lambda calculus with explicit substitutions and graphical equivalence of terms}
